The Unicode Cookbook \linebreak for Linguists

### 2018-03-06 book/chapters/ipa_meets_unicode.tex
updates section name
closes issue #6
closes issue #7
updates IPA reference
these changes almost complete the remaining reviews

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into discrete segments using a Latin-based
alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is
not trivial to encode IPA characters electronically. In this chapter we discuss
various pitfalls with the encoding of IPA in the Unicode Standard. We will
specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0
version of Unicode \citep{Unicode2014}.

@@

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the real retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        bevavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel, which is encoded in IPA with a
``single-story'' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
front unrounded vowel, which is encoded in IPA with a ``double-story'' <a>
\textsc{latin small letter a} at \uni{0061}, basically treating them as
homoglyphs, although they are different phonetic symbols. But even among
linguists this distinction leads to problems. For example, as pointed out by
\citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of
languages are reported to have a five-vowel system \citep{Maddieson1984}.
Historically, linguistic descriptions tend not to include precise audio
recording and measurements of formants, so this may lead one to ask if the many
<a> characters that are used in phonological description reflects a
transcriptional bias. The common use of <a> in transcriptions could be in part
due to the ease of typing the letter on an English keyboard (or for older
descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e., [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

@@

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
complex, there are at least two different Unicode homoglyphs for the low and
high level tones, namely <\diaf{0340}> \textsc{combining grave tone mark} at
\unif{0340} vs.~<\diaf{0300}> \textsc{combining grave accent} at \unif{0300} for
low tone, and <\diaf{0341}> \textsc{combining acute tone mark} at \unif{0341}
vs.<\diaf{0301}> \textsc{combining acute accent} at \unif{0301} for high tone.}
The IPA also proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are much
less often used, but are more consistent for contours.\footnote{Not sanctioned
by the IPA, but nevertheless widely attested, is the usage of superscript
numbers for marking tones.} Tone letters in the IPA have five different levels,
and sequences of these letters can be used to indicate contours. Well-designed
fonts will even merge a sequence of tone letters into a contour. For example,
compare the font Linux Libertine, which does not merge tone letters
<{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which merges this
sequence of four tone letters into a single contour <\charis{˥˨˧˩}>. For strict
IPA encoding we propose to standardize on tone letters.

@@

Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
question whether this issue is at all apparent to the working linguist, or if
they simply use the \uni{0067} because it is easily keyboarded and thus saves
time, whereas the latter must be cumbersomely inserted as a special symbol in
most software. The International Phonetic Association has taken the stance that
both the keyboard \textsc{latin small letter g} and the \textsc{latin small
letter script g} are valid input characters for the voiced velar plosive.
Unfortunately, this decision further introduces ambiguity for linguists trying
to adhere to a strict Unicode Standard IPA encoding. For strict IPA encoding we
propose to standardize on the more idiosyncratic \textsc{latin small letter
script g} at \uni{0261}.

@@

Another pitfall relates to the question of what are diacritics. The problem is that
the meaning of the term diacritics as used by the IPA is not the same as is used
in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
definition combined with the character before them (to form so-called default
grapheme clusters, see Section~\ref{the-unicode-approach}). In contrast, Spacing
Modifier Letters are by definition \emph{not} combined into grapheme clusters
with the preceding character, but simply treated as separate letters. In the
context of the IPA, the following IPA-diacritics are actually Spacing Modifier
Letters in the Unicode Standard:

@@

	\item[] Tone letters, like: 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˨> \textsc{modifier letter low tone bar} at \uni{02E8}
	  \item[] and others like this
	\end{itemize}

	\item[] Superscript letters, like:
	\begin{itemize}
	  \item[] <\dia{02B0}> \textsc{modifier letter small h} at \uni{02B0}
	  \item[] <\dia{02E4}> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
	  \item[] <\dia{207F}> \textsc{superscript latin small letter n} at \uni{207F}
	  \item[] and many more like this
	\end{itemize}

@@

   \item According to the IPA, <ʰ> can be used both as <\dia{02B0}> for
         post-aspiration (following the base character) and as
         <\diareverse{02B0}> for pre-aspiration (preceding the base
         character), so there is no consistent direction in which this
         diactritic should bind. Note that there is yet another homoglyph,
         namely the \textsc{combining latin small letter h} at \uni{036A}.
         We propose not to use this combining diacritical mark, but to
         standardize on Unicode modifier letters for all superscript
         letters in IPA.

@@

The most controversial issue regarding symbols debated at the convention was the
representation for voiceless implosives \citep[62]{Ladefoged1990a}. In
accordance with the principles of the IPA, as outlined in Section
\ref{the-international-phonetic-alphabet}, distinct symbols are favoured for
cases of phonological contrast. Further, also convenience of display in the chart
must be taken into account when arguing for or against the inclusion or deletion
of IPA symbols in the IPA chart. Finally, the inclusion or deletion of symbols
should consider the current state of phonetic knowledge of the world's
languages.

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode Standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

@@

% ==========================
\section{Recommendations}
\label{ipa-recommendations}
% ==========================

@@

Summarizing the pitfalls as discussed in this chapter, we propose to define
three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@
Informally speaking, valid-IPA represents the current state of the IPA
\citep{IPA2005}. Strict-IPA represents a more constrained version of IPA, while
widened-IPA is a slightly extended version of IPA, allowing a few more symbols.

>>>>

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into discrete segments using a Latin-based
alphabet. Although IPA is reasonably easily adhered to with pen and paper, it is
not trivial to encode IPA characters electronically. In this chapter we discuss
various pitfalls with the encoding of IPA in the Unicode Standard. We will
specifically refer to the 2015 version of the IPA \citep{IPA2015} and the 10.0.0 version of Unicode \citep{Unicode2017}.

@@

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the real retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.        
  \item There are two mostly identical-looking Unicode characters for the superscript
        <ʰ>: the \textsc{combining latin small letter h} at \uni{036A} and the
        \textsc{modifier letter small h} at \uni{02B0}. Making the situation 
        even more problematic is that they have different behavior (see 
        Section~\ref{pitfall-different-notions-of-diacritics}). To harmonize the 
        behavior of <ʰ> with other superscript letters, we propose to 
        standardize on the modifier letter at \uni{02B0} (see 
        Section~\ref{pitfall-no-unique-diacritic-ordering}).

@@

Conversely, non-linguists are unlikely to distinguish any semantic difference
between an open back unrounded vowel, which is encoded in IPA with a
``single-story'' <ɑ> \textsc{latin small letter alpha} at \uni{0251}, and the open
front unrounded vowel, which is encoded in IPA with a ``double-story'' <a>
\textsc{latin small letter a} at \uni{0061}, basically treating them as
homoglyphs, although they are different phonetic symbols. But even among
linguists this distinction leads to problems. For example, as pointed out by
\citet{Mielke2009}, there is a problem stemming from the fact that about 75\% of
languages are reported to have a five-vowel system \citep{Maddieson1984}.
Historically, linguistic descriptions tend not to include precise audio
recording and measurements of formants, so this may lead one to ask if the many
<a> characters that are used in phonological description reflects a
transcriptional bias. The common use of <a> in transcriptions could be in part
due to the ease of typing the letter on an English keyboard (or for older
descriptions, the typewriter). We found it to be exceedingly rare that a
linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts. For
instance, uvular and pharyngeal consonants only occur around /a/ and /o/. Hence,
the characterization of /a/ and /o/ as a natural class of (i.e.\ [+back]
vowels), as opposed to /i/ and /e/.''} They simply use <a> as long as there is
no opposition to <ɑ>.

@@

Lastly, before we move on from the pitfall of IPA homoglyphs in Unicode to the pitfall of 
homoglyphs in IPA, we are aware of one example that illustrates both pitfalls. 
Consider for example what one reviewer coined \textsc{i dot-suppression}. 
When combining, say the \textsc{latin small letter i} <i> at \uni{0069} with 
the \textsc{combining acute accent} <\dia{0301}> at \uni{0301}, the result is 
 the combination of these two characters into <í> or the associated 
precomposed form \textsc{latin small letter i with acute} <í> at \uni{00ED}. 
Typographically, the accent mark replaces the dot. In IPA, the <í> denotes 
a high front unrounded vowel with high tone. However, the result of losing 
the dot makes this IPA symbol look very similar to the near-high near-front 
unrounded vowel <ɪ>, when it also has the high tone marker: <ɪ́>. To boot, 
when an accent mark is added to \textsc{latin small letter i with stroke} <ɨ> at 
\uni{0268}, the dot is not suppressed but retained, i.e. <ɨ́>.\footnote{According 
to the Unicode Standard, \textsc{latin small letter i with stroke} <ɨ> at 
U+0268 cannot be decomposed into, say, \textsc{latin small letter i} <i> at 
U+0069 and \textsc{combining short stroke overlay} <\dia{0335}> at U+0335. 
We discuss the pitfall of missing decomposition forms 
in Section \ref{pitfall-missing-decomposition}.}

@@

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone (cf.\ \cite{Maddieson1990}). There is an old tradition 
to use diacritics on vowels to mark different tone levels \citep{IPA1949}. Prior to the 1989 Kiel 
convention, IPA-approved tone symbols included diacritics above or below the vowel or syllable, 
e.g.\ high and low tones marked with macrons (<\diaf{0304}>, <\diaf{0331}>), and acute and grave 
accents for high rising tone <\diaf{0301}>, low rising tone <\diaf{0317}>, high falling tone 
<\diaf{0300}> and low falling tone <\diaf{0316}>. These tone symbols, however,  
had failed to catch on (probably) due to aesthetic objections and matters of adequacy for 
transcription \citep[29]{Maddieson1990}.

After the 1989 Kiel convention, the accent tone symbols were updated to the tradition 
that we are familiar with today and which was already in wide use by Africanists and others, 
namely level tones <\diaf{030B}, \diaf{0301}, \diaf{0304}, \diaf{0300}, \diaf{030F}> 
and contour tones <\diaf{030C}, \diaf{0302}, \diaf{1DC4}, \diaf{1DC5}, \diaf{1DC8}>.\footnote{To make things even more
complicated, there are at least two different Unicode homoglyphs for the low and
high level tones, namely <\diaf{0340}> \textsc{combining grave tone mark} at
\unif{0340} vs.~<\diaf{0300}> \textsc{combining grave accent} at \unif{0300} for
low tone, and <\diaf{0341}> \textsc{combining acute tone mark} at \unif{0341}
vs.<\diaf{0301}> \textsc{combining acute accent} at \unif{0301} for high tone.} 
In addition, the IPA also adopted tone letters developed by \citet{Chao1930}, e.g. <˥˦˧˨˩>, 
which were in wide use by Asianists.\footnote{Not sanctioned
by the IPA, but nevertheless widely attested, is the use of superscript
numbers for marking tones, also proposed by \citet{Chao1930}. One issue to note here is 
that superscript numbers can be either regular numbers that are formatted as 
superscript with a text processor, or they can be separate superscript characters, 
as defined in the Unicode Standard (see: \url{https://en.wikipedia.org/wiki/Superscripts_and_Subscripts}). 
This divide means that searching text is dependent 
on how the author formatted or encoded the superscript numbers.} 
Tone letters in the IPA have five different levels,
and sequences of these letters can be used to indicate contours. Well-designed
fonts will even merge a sequence of tone letters into a contour. For example,
compare the font Linux Libertine, which does not merge tone letters
<{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which merges this
sequence of four tone letters into a single contour <\charis{˥˨˧˩}>. For strict
IPA encoding we propose to standardize on tone letters.

@@

Second, we commonly encounter the use of <g> \textsc{latin small letter g} at
\uni{0067}, instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to
question whether this issue is at all apparent to the working linguist, or if
they simply use the \uni{0067} because it is easily keyboarded and thus saves
time, whereas the latter must be cumbersomely inserted as a special symbol in
most software. The International Phonetic Association has taken the stance that
both the keyboard \textsc{latin small letter g} and the \textsc{latin small
letter script g} are valid input characters for the voiced velar plosive \citep[19]{IPA1999}.
Unfortunately, this decision further introduces ambiguity for linguists trying
to adhere to a strict Unicode Standard IPA encoding. For strict IPA encoding we
propose to standardize on the more idiosyncratic \textsc{latin small letter
script g} at \uni{0261}.

@@

Another pitfall relates to the question of what are diacritics. The problem is that
the meaning of the term diacritics as used by the IPA is not the same as is used
in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
definition combined with the character before them (to form so-called default
grapheme clusters, see Chapter~\ref{the-unicode-approach}). In contrast, Spacing
Modifier Letters are by definition \emph{not} combined into grapheme clusters
with the preceding character, but simply treated as separate letters. In the
context of the IPA, the following IPA-diacritics are actually Spacing Modifier
Letters in the Unicode Standard:

@@

	\item[] Tone letters, including but nut limited to: 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˨> \textsc{modifier letter low tone bar} at \uni{02E8}
	\end{itemize}

	\item[] Superscript letters, including but not limited to:\footnote{The Unicode Standard defines the 
	  \textit{Phonetic Extensions} block that defines symbols used in phonetic notation 
	  systems, including the Uralic Phonetic Alphabet, Americanist and Russianist phonetic notations, 
	  Oxford English and American dictionaries, etc. Among other symbols, the 
	  \textit{Phonetic Extensions} block includes 
	  the superscript letters <\textsuperscript{m, ŋ, b}>, which are not valid 
	  IPA characters, although we have seen them used in linguistic practice.}
	\begin{itemize}
	  \item[] <\dia{02B0}> \textsc{modifier letter small h} at \uni{02B0}
	  \item[] <\dia{02E4}> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
	  \item[] <\dia{207F}> \textsc{superscript latin small letter n} at \uni{207F}
	\end{itemize}

@@

   \item According to the IPA, <ʰ> can be used both as <\dia{02B0}> for
         post-aspiration (following the base character) and as
         <\diareverse{02B0}> for pre-aspiration (preceding the base
         character), so there is no consistent direction in which this
         diacritic should bind. Note that there is yet another homoglyph,
         namely the \textsc{combining latin small letter h} at \uni{036A}.
         We propose not to use this combining diacritical mark, but to
         standardize on Unicode modifier letters for all superscript
         letters in IPA.

@@

The most controversial issue regarding symbols debated at the convention was the
representation for voiceless implosives \citep[62]{Ladefoged1990a}. In
accordance with the principles of the IPA, as outlined in Section
\ref{the-international-phonetic-alphabet}, distinct symbols are favored for
cases of phonological contrast. Further, also convenience of display in the chart
must be taken into account when arguing for or against the inclusion or deletion
of IPA symbols in the IPA chart. Finally, the inclusion or deletion of symbols
should consider the current state of phonetic knowledge of the world's
languages.

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode Standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.\ the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

@@

% ==========================
\section{Unicode IPA Recommendations}
\label{ipa-recommendations}
% ==========================

@@

Summarizing the pitfalls as discussed in this chapter, we propose to define
three different IPA encodings: strict-IPA, valid-IPA and widened-IPA.\@
Informally speaking, valid-IPA represents the current state of the IPA
\citep{IPA2015}. Strict-IPA represents a more constrained version of IPA, while
widened-IPA is a slightly extended version of IPA, allowing a few more symbols.
